---
title: "Houston Winter Storm and Power Outage Analysis"
author: Ava Robillard
date: 10/30/25
format: html
execute:
  message: false
  warning: false
---

As climate change increases the frequency of extreme weather events, the impacts associated with these events are felt more intensely by the communities that live there. In February of 2021, Texas experienced a power crisis due to multiple severe winter storms. By understanding how the impact of losing power was felt across the Houston community, we can allocate resources and plan more strategically for future weather events.

**Objective**: Use VIIRS, OpenStreetMap, and U.S. Census data to visualize the impacts of the winter storms by estimating the number of homes in the Houston metropolitan area that lost power and combine this with socioeconomic data about median household income to investigate whether these impacts were disproportionately felt.

## Load packages

```{r}
library(tidyverse)
library(stars)
library(sf)
library(terra)
library(tmap)
```

1.  Load in all data

```{r}
# Read in night light data tiles
tile1_2021_02_07 <- stars::read_stars(here::here("data", "VNP46A1", "VNP46A1.A2021038.h08v05.001.2021039064328.tif"))

tile2_2021_02_07 <- stars::read_stars(here::here("data", "VNP46A1", "VNP46A1.A2021038.h08v06.001.2021039064329.tif"))

tile1_2021_02_16 <- stars::read_stars(here::here("data", "VNP46A1", "VNP46A1.A2021047.h08v05.001.2021048091106.tif"))

tile2_2021_02_16 <- stars::read_stars(here::here("data", "VNP46A1", "VNP46A1.A2021047.h08v06.001.2021048091105.tif"))

# Combine tiles for each date into a single continuous raster- maps
lights_2021_02_07 <- st_mosaic(tile1_2021_02_07, tile2_2021_02_07)
  
lights_2021_02_16 <- st_mosaic(tile1_2021_02_16, tile2_2021_02_16)

# Crop each light raster to Houston area
bb <- st_bbox(c(xmin = -96.5, xmax = -94.5, ymax = 30.5, ymin = 29.0), crs = st_crs(lights_2021_02_07))
lights_2021_02_07 <- st_crop(lights_2021_02_07, bb)

lights_2021_02_16 <- st_crop(lights_2021_02_16, bb)

# do i want to transform crs here? right now it is EPSG 4326
```

Map light differences

```{r}
# Plot before and after- pick which style and fix legend 
before_lights <- tm_shape(lights_2021_02_07) +
  tm_raster(col.scale = tmap::tm_scale_continuous(
    values = "-brewer.Greys"),
    col.legend = tmap::tm_legend(
      title = "Luminance (cm-2sr-1)",
      position = tmap::tm_pos_in("left", "bottom"))
    ) +
  tm_title(text = "Houston Area Luminance: Feb 7th, 2021") +
  tmap::tm_compass() 

after_lights <- tm_shape(lights_2021_02_16) +
  tm_raster(palette = colorRampPalette(c("black", "white")) (6),
      title = "Luminance (cm-2sr-1)"
    ) +
  tm_title(text = "Houston Area Luminance: Feb 16th, 2021") +
  tmap::tm_compass(position = c("right", "bottom")) 

tmap_arrange(before_lights, after_lights)
```

```{r}
#| message: false
# Read in roads data and filter to only class motorway
roads <- sf::st_read(here::here("data/gis_osm_roads_free_1.gpkg"),
                     query = "SELECT * FROM gis_osm_roads_free_1 WHERE fclass='motorway'",
                     quiet = TRUE)

# Transform CRS to NAD83 / Texas Centric Albers Equal Area
roads <- st_transform(roads, 3083)

# Read in buildings data and filter to only types of houses
houses <- sf::st_read(here::here("data/gis_osm_buildings_a_free_1.gpkg"),
                      query = "SELECT * FROM gis_osm_buildings_a_free_1 WHERE (type IS NULL AND name IS NULL)
OR type in ('residential', 'apartments', 'house', 'static_caravan', 'detached')",
                      quiet = TRUE)

# Transform CRS to NAD83 / Texas Centric Albers Equal Area
houses <- st_transform(houses, 3083)
```

```{r}
# Read in entire geodatabase
socioeconomic <- st_read(here::here("data/ACS_2019_5YR_TRACT_48_TEXAS.gdb"), quiet = TRUE)

# Explore socioeconomic geodatabase layers
layer_info <- st_layers(dsn = here::here("data/ACS_2019_5YR_TRACT_48_TEXAS.gdb"))
#layer_info

# Create geometry layer using appropriate layer names
ACS_tract_geometry <- st_read(here::here("data", "ACS_2019_5YR_TRACT_48_TEXAS.gdb"),
                              layer = "ACS_2019_5YR_TRACT_48_TEXAS",
                              quiet = TRUE)

# Create attribute layer 
ACS_tract_attributes <- st_read(here::here("data", "ACS_2019_5YR_TRACT_48_TEXAS.gdb"),
                                layer = "X19_INCOME",
                                quiet = TRUE)

# Rename GEOID_Data column in ACS_tract_geometry to GEOID for a correct join
ACS_tract_geometry <- ACS_tract_geometry %>% 
  rename(GEOID_old = GEOID) %>% 
  rename(GEOID = GEOID_Data)

# Combine layers and reproject to NAD83 / Texas Centric Albers Equal Area
ACS_tracts <- ACS_tract_geometry %>% 
  left_join(ACS_tract_attributes, by = "GEOID") %>% 
  st_as_sf() 

# Crop to Houston area
bb <- st_bbox(c(xmin = -96.5, xmax = -94.5, ymax = 30.5, ymin = 29.0), crs = st_crs(ACS_tracts))
ACS_tracts <- st_crop(ACS_tracts, bb) 

# Transform CRS to NAD83 / Texas Centric Albers Equal Area
ACS_tracts <- st_transform(ACS_tracts, 3083)
```

```{r}
## FIX THIS!!!
# Ensure data sets have same coordinate reference systems 
if(crs(roads) == crs(houses)) {
  print("Coordinate reference systems match")
} else{
  warning("Updating coordinate reference systems to match")
  # transform data to match
  boundary <- st_transform(roads, st_crs(houses))
}

if(st_crs(lights_2021_02_07) == st_crs(lights_2021_02_16)) {
  print("Coordinate reference systems match")
} else{
  warning("Updating coordinate reference systems to match")
  # transform data to match
  boundary <- st_transform(lights_2021_02_07, st_crs(lights_2021_02_16))
}

if(st_crs(lights_2021_02_07) == crs(roads)) {
  print("Coordinate reference systems match")
} else{
  warning("Updating coordinate reference systems to match")
  # transform data to match
  boundary <- st_transform(lights_2021_02_07, st_crs(roads))
}

if(st_crs(lights_2021_02_07) == st_crs(ACS_tracts)) {
  print("Coordinate reference systems match")
} else{
  warning("Updating coordinate reference systems to match")
  # transform data to match
  boundary <- st_transform(ACS_tracts, st_crs(lights_2021_02_07))
}
```

2.  Find locations that experienced a blackout by creating a mask

```{r}
# Find the change in night lights intensity pre and post storm
# Check for same extent and resolution first!
light_difference <- lights_2021_02_07 - lights_2021_02_16

# Assign NA to cells that experienced a change less than 200 nW cm-2sr-1
light_difference[light_difference < 200] <- NA

# Vectorize the blackout mask and fix invalid geometries
light_diff_vec <- light_difference %>% 
  st_as_sf() %>% #add st_is_valid() in here before st_make_valid, then check output
  st_make_valid() %>% 
  st_transform(3083)
```

3.  Exclude highways from analysis

```{r}
#| warning: false
# Check units of CRS for distance argument in buffer
units <- st_crs(roads)$units

# Create 200m buffer around all highways
highway_buffer <- st_buffer(roads, dist = 200)

# Combine buffer geometries- dissolve multi-polygons
highway_union <- st_union(highway_buffer) 

# Create data frame for blackouts that are not in the highway union
true_blackouts <- st_difference(light_diff_vec, highway_union)

# Visualize buffer- 200m is small!- is this properly filtering?
tm_shape(highway_union) +
  tm_polygons(col = "black") +
  tm_shape(roads) +
  tm_lines(col = "pink", lwd = 0.4) +
  tm_title(text= "200m highway buffer") +
tm_shape(true_blackouts) +
  tm_polygons(col = "lightblue")
```

The units of the roads CRS is `r units`.

4.  Identify homes that experienced blackouts by combining the locations of homes and blackouts

```{r}
#| warning: false
# Filter homes to those that intersect with the true blackouts polygon
blackout_houses <- st_intersection(houses, true_blackouts)

# Get number of affected homes by number of rows
n_affected_houses <- nrow(blackout_houses)

### FIX MAP HERE
tm_shape(blackout_houses) +
  tm_polygons() +
  tm_title(text = "Houses that experienced blackouts")
```
The estimated number of homes in Houston that lost power between February 7th and February 16th, 2021 is `r n_affected_houses`.

5.  Identify the census tracts likely impacted by the blackout

```{r}
# In socioeconomic ACS_tracts data, make a column with true where there is a home with a blackout (blackout_houses) in the census tract- then join with full to get falses-

# Intersect ACS_tracts data and blackout homes data- makes all true?
blackedout_tracts <- st_filter(ACS_tracts, blackout_houses) %>% 
  mutate(blackout = TRUE)

# TRUE/FALSE column
census <- ACS_tracts %>% 
  mutate(blackout = ACS_tracts$GEOID %in% blackedout_tracts$GEOID)

# Filter census data frame to reduce computation time
census <- census %>% 
  select('GEOID', 'B19013e1', 'Shape', 'TRACTCE', 'blackout')

# Select true blackouts for plotting
census_true <- census %>% 
  filter(blackout == TRUE)

# plot full census tracts and overlay with plotted true census tracts
tm_shape(census) +
  tm_polygons() +
tm_shape(census_true) +
  tm_polygons(fill = 'blue') 
```

6.  Create plot comparing distributions of median household income for census tracts that did and did not experience blackouts

```{r}
#| warning: false
#| message: false
median_income_hist <- ggplot(census, aes(x = B19013e1)) +
  geom_histogram(fill = 'darkblue') +
  facet_wrap(~blackout)

median_income_hist
```
